---
title: '7.9'
author: "Bingyan Liang"
date: "2023-03-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fit a proportional odds model
We fit a proportional odds model to the data with the five risk factors as covariates. The results are shown in Table 7.1. It can be seen that younger
age, male sex, histories of needle sharing, imprisonment, and drug injection in jail all increase the risk of HIV sero-conversion. 

```{r, echo=FALSE}
library(survival)
library(IntCens)

## BMA HIV study
## Bangkok Metropolitan Administration HIV Study
data <- read.table("bam.txt")


# get the response data for ICSurvICM()
delta <- data$delta
gamma <- data$gamma
n <- nrow(data)
U <- data$U
V <- data$V


# Fit a proportional odds model
obj.PO <- ICSurvICM(delta,gamma,U,V,Z=data[,5:9],model="PO")
print.ICSurv(obj.PO)
```


```{r, echo=FALSE}
#regression parameter
beta <- obj.PO$beta
se <- sqrt(diag(obj.PO$var))
c1 <- round(exp(beta),2)
c2 <- paste0("(",round(exp(beta-1.96*se),2),", ",
             round(exp(beta+1.96*se),2),")")
```
```{r}
noquote(cbind(c1,c2))
```

## Proportional hazard model in 7.2.4

```{r, echo=FALSE}
library(survival)
library(IntCens)

## BMA HIV study
## Bangkok Metropolitan Administration HIV Study
data <- read.table("bam.txt")


# get the response data for ICSurvICM()
delta <- data$delta
gamma <- data$gamma
n <- nrow(data)
U <- data$U
V <- data$V


# Fit a proportional hazards model
obj.PH <- ICSurvICM(delta,gamma,U,V,Z=data[,5:9],model="PH")
print.ICSurv(obj.PH)

```


```{r, echo=FALSE}
# Table 7.1
# construct a table for hazard ratio and
# 95% confidence intervals

#regression parameter
beta <- obj.PH$beta
se <- sqrt(diag(obj.PH$var))
c1 <- round(exp(beta),2)
c2 <- paste0("(",round(exp(beta-1.96*se),2),", ",
   round(exp(beta+1.96*se),2),")")
```
```{r}
noquote(cbind(c1,c2))
```

Compared with 7.2.4, the effect of sex in the odds model is highly significant (p value 0.1) -- subjects who is male are 1.65 times as likely to become HIV sero-positive than female. Among all the factors, the effect of sex, need and inject are larger than them in proportional hazard model

