---
title: "midterm"
author: "Bingyan Liang"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Descriptive statistics:

Summarize the demographic variables (gender, age, race), BMI, and medical history(LVEF, MI, HF, diabetes, etc.) by treatment and overall. Be sure to also calculate the (composite) event rate of death/hospitalization.

```{r include=FALSE}
library(survival)
library(table1)
library(knitr)

data <- read.table('invested1.txt')
df <- data
## table 1

## label
data <- na.omit(data)
data$status <- factor(data$status,
                      levels = c(0,1),
                      labels = c("censoring","death/hospitalization"))

# original tableone
table1 <- table1(~ status + gender+ age + race + dbmi + lveflt40 + priormi + priorhf + diab + renal + ischstr + pad | trtmnt, data = data )
print(table1)
table <- data.frame(table1)

# Numerator: total # of events
# (N of events is sum of status variable)
num.D <- c(sum(df$status[df$trtmnt=='HD']), 
           sum(df$status[df$trtmnt=='SD']), 
           sum(df$status))

# Denominator: total length of follow-up (year)
denom.D <- c(sum(df$time[df$trtmnt=='HD']), 
             sum(df$time[df$trtmnt=='SD']), 
             sum(df$time))/12

# Death rate
rate1 <- round(num.D/denom.D,3)
rate <- c("death/hospitalization rate(1/year)", rate1)

# add a row to table 1
tableone <- rbind(table, rate)

tableone[9,1] <- "age(years)"
```
```{r}
kable(tableone)

```

## 2.Graphical analysis of subgroup effect:

Plot the treatment-specific (HD vs SD) Kaplan—Meier curves for hospitalization-free survival probabilities within each gender and each race. 
Comment on the difference in treatment effect across genders and races.

First we plot the overall K-M curves.

```{r, echo=FALSE}
par(mfrow=c(1,1.5))
df <- na.omit(df)
## overall
obj <- survfit(Surv(time,status)~trtmnt,data=df)
plot(obj, ylim = c(0,1), lwd=2,frame=F, lty=c(2,1),
     xlab="Time (months)",ylab="hospitalization-free survival probabilities",main="Overall",
     cex.lab=1.5,cex.axis=1.5,cex.main=1.5)
legend('bottomleft', lty=2:1,c("HD","SD"),
       lwd=2,cex=1.5)
```

Next, we plot the event-free survival probabilities within each race.

```{r, echo=FALSE}
## black
obj.bl <- survfit(Surv(time,status)~trtmnt,data=df[df$race=='Black/African American',])
plot(obj.bl, lwd=2,frame=F, lty=c(2,1),
     xlab="Time (months)",ylab="hospitalization-free survival probabilities",main="Black/African American",
     cex.lab=1.5,cex.axis=1.5,cex.main=1.5)
legend('bottomleft',lty=2:1,c("HD","SD"),
       lwd=2,cex=1.5)

## white
obj.wh<- survfit(Surv(time,status)~trtmnt,data=df[df$race=='White',])
plot(obj.wh, lwd=2,frame=F, lty=c(2,1),
     xlab="Time (months)",ylab="hospitalization-free survival probabilities",main="White",
     cex.lab=1.5,cex.axis=1.5,cex.main=1.5)
legend('bottomleft',lty=2:1,c("HD","SD"),
       lwd=2,cex=1.5)

## Aboriginal/Native American
obj.ab<- survfit(Surv(time,status)~trtmnt,data=df[df$race=='Aboriginal/Native American',])
plot(obj.ab, lwd=2,frame=F, lty=c(2,1),
     xlab="Time (months)",ylab="hospitalization-free survival probabilities",main="Aboriginal/Native American",
     cex.lab=1.5,cex.axis=1.5,cex.main=1.5)
legend('bottomleft',lty=2:1,c("HD","SD"),
       lwd=2,cex=1.5)

## Asian/Pacific Islander
obj.as<- survfit(Surv(time,status)~trtmnt,data=df[df$race=='Asian/Pacific Islander',])
plot(obj.as, lwd=2,frame=F, lty=c(2,1),
     xlab="Time (months)",ylab="hospitalization-free survival probabilities",main="Asian/Pacific Islander",
     cex.lab=1.5,cex.axis=1.5,cex.main=1.5)
legend('bottomleft',lty=2:1,c("HD","SD"),
       lwd=2,cex=1.5)

## other
obj.ot<- survfit(Surv(time,status)~trtmnt,data=df[df$race=='Other',])
plot(obj.ot, lwd=2,frame=F, lty=c(2,1),
     xlab="Time (months)",ylab="hospitalization-free survival probabilities",main="Other",
     cex.lab=1.5,cex.axis=1.5,cex.main=1.5)
legend('bottomleft',lty=2:1,c("HD","SD"),
       lwd=2,cex=1.5)
```

Then, plot within each gender.

```{r, echo=FALSE}
## male
obj.m <- survfit(Surv(time,status)~trtmnt,data=df[df$gender=='Male',])
plot(obj.m, lwd=2,frame=F, lty=c(2,1),
     xlab="Time (months)",ylab="hospitalization-free survival probabilities",main="Male",
     cex.lab=1.5,cex.axis=1.5,cex.main=1.5)
legend('bottomleft',lty=2:1,c("HD","SD"),
       lwd=2,cex=1.5)

## female
obj.f<- survfit(Surv(time,status)~trtmnt,data=df[df$gender=='Female',])
plot(obj.f, lwd=2,frame=F, lty=c(2,1),
     xlab="Time (months)",ylab="hospitalization-free survival probabilities",main="Female",
     cex.lab=1.5,cex.axis=1.5,cex.main=1.5)
legend('bottomleft',lty=2:1,c("HD","SD"),
       lwd=2,cex=1.5)
```

From the plots, we can conclude that, within different races, in Black/African American and Aboriginal/Native American groups, the HD group appears to enjoy higher event-free survival probabilities compared to the SD group. But overall, there did not appear to have a significant difference between the two treatment groups by race. To conduct a formal test on the effect of doses, we use the standard log-rank test stratified by race:
```{r}
## Stratified log-rank test (by race)
survdiff(Surv(time,status)~trtmnt+ strata(race), data=df)
```
The result support our conclusion that there is no significant difference between two treatment groups.

Also, within different genders, we can tell from the plot that there is no significant difference between two treatment groups, and the standard log-rank test stratified by gender can support our conclusion:
```{r}
## Stratified log-rank test (by gender)
survdiff(Surv(time,status)~trtmnt+ strata(gender), data=df)
```

## 3. Multiple regression analysis:
Build a Cox proportional hazards regression model to analyze the treatment effect and other risk factors for death/hospitalization. 
Please include the following steps:
    a. Assess formally whether treatment effect differs significantly by gender or race (by testing interactions);
    b. Check the proportionality assumption on the each of the covariates and take steps to address possible violations;
    c. Check whether the linear form is appropriate for age and BMI; make proper transformation/categorization in case of non-linearity;
    d. Interpret the analysis results of your final model, e.g., what risk factors are most strongly or significantly associated with the primary endpoint and to what degree.

First, conduct the cox proportional hazard model.

```{r, echo=FALSE}
#fit the Cox proportional hazards model
obj <- coxph(Surv(time,status)~age+factor(gender)+factor(race)+dbmi+factor(lveflt40)+factor(priormi)
             +factor(priorhf)+factor(diab)+factor(renal)+factor(ischstr)+factor(pad),data=df)
summary(obj)

#fit the Cox proportional hazards model with interaction terms
obj.gen <- coxph(Surv(time,status)~factor(gender),data=df)
summary(obj.gen)
obj.race<- coxph(Surv(time,status)~factor(race),data=df)
summary(obj.race)
```

The tests shows that for genders, there is no significance. And for race, it can give significant contributions on the result.

We can check the proportionality assumption on the each of the covariates by assess a residual analysis.

```{r, echo=FALSE}
#Schoelfeld residuals
#produce proportionality test results
sch <- cox.zph(obj) 
print(sch) 
#Plot scaled Schoelfeld residuals for each covariate
par(par(mar = c(4, 4, 2, 2)), mfrow=c(3,3))
plot(sch,xlab="Time (months)",lwd=2,cex.lab=1.2,cex.axis=1.2)
```

From both the graphical and numerical results, it seems that proportionality is problematic on age, gender and Prior myocardial infarction.
Since gender and priormi is categorical, we can take them off from the covariate list and use them instead as stratifying variables.
First we remove gender:
```{r, echo=FALSE}
#fit the Cox proportional hazards model
obj.stra <- coxph(Surv(time,status)~age+factor(race)+dbmi+factor(lveflt40)+factor(priorhf)+factor(diab)
                  +factor(renal)+factor(ischstr)+factor(pad)+strata(gender)+factor(priormi),data=df)
sch.stra <- cox.zph(obj.stra)
print(sch.stra)
```
We can see that the global proportionality test goes to non-significant.
Now we also remove the Prior myocardial infarction:
```{r, echo=FALSE}
#fit the Cox proportional hazards model
obj.stra <- coxph(Surv(time,status)~age+factor(race)+dbmi+factor(lveflt40)+factor(priorhf)+factor(diab)
                  +factor(renal)+factor(ischstr)+factor(pad)+strata(gender)+strata(priormi),data=df)
sch.stra <- cox.zph(obj.stra)
print(sch.stra)
```
We see from the above output that the global proportionality test becomes non-significant after removing gender and Prior myocardial infarction.

To check whether the linear form is appropriate for age and BMI, we generate scatter plots for the martingale residuals against the covariates overlaid with a smoothed curve to better capture the trend.
```{r, echo=FALSE}
## Martingale residuals
mart_resid <- resid(obj,type='martingale')

#plot the martingale residuals against age and bmi

## Age
par(mfrow=c(1,2))
plot(df$age, mart_resid,
     xlab="age (years)", ylab="Martingale residuals",
     main='age',cex.lab=1.5)
lines(lowess(df$age, mart_resid),lwd=2)
abline(0,0,lty=3,lwd=2)

## bmi
plot(df$dbmi, mart_resid,
     xlab="bmi", ylab="Martingale residuals",
     main='bmi',cex.lab=1.2)
lines(lowess(df$dbmi, mart_resid),lwd=2)
abline(0,0,lty=3,lwd=2)

## log-Age
par(mfrow=c(1,2))
plot(log(df$age), mart_resid,
     xlab="log-age", ylab="Martingale residuals",
     main='log-age',cex.lab=1.5)
lines(lowess(log(df$age), mart_resid),lwd=2)
abline(0,0,lty=3,lwd=2)

## log-bmi
plot(log(df$dbmi), mart_resid,
     xlab="log-bmi", ylab="Martingale residuals",
     main='log-bmi',cex.lab=1.2)
lines(lowess(log(df$dbmi), mart_resid),lwd=2)
abline(0,0,lty=3,lwd=2)
```

Linear age and bmi are both problematic: there are substantial excess failures in elder people and people with higher bmi. To address it, we group age into a categorical variable agec, with agec=1: ≤ 40 years; agec=2: (40, 60] years, agec=3: (60, 80] years, and agec=4: > 80 years. Group bmi into a categorical variable bmic, with bmic=1: ≤ 30; bmic=2: (30, 50], bmic=3: (50, 70], and agec=4: > 70. The model is re-fitted, whose results are summarized below.

```{r, echo=FALSE}
# To address non-linear age
# categorize age in agec
# age<=40: agec=1
# 40<age<=60: agec=2
# 60<age<=80: agec=3
# age>80: agec=4
df$agec <- (df$age<=40)+2*(df$age>40&df$age<=60)+3*(df$age>60&df$age<=80)+4*(df$age>80)
df$bmic <- (df$dbmi<=30)+2*(df$dbmi>30&df$dbmi<=50)+3*(df$dbmi>50&df$dbmi<=70)+4*(df$dbmi>70)

#re-fit the model with agec and bmic
obj.stra.final <- coxph(Surv(time,status)~factor(agec)+factor(race)+factor(bmic)+factor(lveflt40)+factor(priorhf)+factor(diab)+factor(renal)+factor(ischstr)+factor(pad)+strata(gender)+strata(priormi),data=df)

#plot the estimated HRs for agec (Figure 4.6)
# and confidence intervals
final.sum <- summary(obj.stra.final)
final.sum

# Plot the age-group-specific HR and confidence
# intervals from the re-fitted model
ci.table=final.sum$conf.int
hr=ci.table[3:5,1]
hr.low=ci.table[3:5,3]
hr.up=ci.table[3:5,4]

par(mfrow=c(1,1))
plot(1:4,c(1,hr),ylim=c(0,2.5),frame=F,xaxt='n',
     xlab="Age (years)", ylab="Hazard ratio",pch=19,cex=1.5,cex.lab=1.2,
     cex.axis=1.2)
axis(1, at=c(1,2,3,4),labels=c("(20, 40]","(40, 60]","(60, 80]","(80, 100]"),cex.axis=1.5)
# horizontal error bars
arrows(2:4, hr.low, 2:4, hr.up, length=0.05, angle=90, code=3, lwd=2)
lines(1:4,c(1,hr),lty=3,lwd=2)

# Plot the bmi-group-specific HR and confidence
# intervals from the re-fitted model

hr1=ci.table[10:12,1]
hr1.low=ci.table[10:12,3]
hr1.up=ci.table[10:12,4]

par(mfrow=c(1,1))
plot(1:4,c(1,hr1),ylim=c(0,5),frame=F,xaxt='n',
     xlab="bmi", ylab="Hazard ratio",pch=19,cex=1.5,cex.lab=1.2,
     cex.axis=1.2)
axis(1, at=c(1,2,3,4),labels=c("(10, 30]","(30, 50]","(50, 70]","(70, 90]"),cex.axis=1.2)
# horizontal error bars
arrows(2:4, hr1.low, 2:4, hr1.up, length=0.05, angle=90, code=3,lwd=2)
lines(1:4,c(1,hr1),lty=3,lwd=2)
```

It can be seen that the age effect and bmi effecr are both nonlinear, with 80-100 years people at highest risk, 60-80 years at lowest risk, and younger people 20-40 years and 40-60 years at a slightly lower risk than older people. 30-50 bmi people at the highest risk and 10-30 bmi with the lowest risk.

In our final model, among all the factor covariates we could see several terms that have significant effects on the result, with p-values:

factor(bmic)3                      8.76e-05 ***

factor(lveflt40)Yes                8.68e-09 ***

factor(priorhf)Yes                 < 2e-16 ***

factor(diab)Yes                    9.67e-07 ***

factor(renal)Yes                   < 2e-16 ***

factor(ischstr)Yes                 0.000116 ***


## 4. Reporting: 
This study compared the effects of high-dose versus standard-dose influenza vaccine on morbidity and mortality in high-risk cardiovascular patients. The results showed that, overall, there was no significant difference in morbidity and mortality between the two doses of vaccine in the high-risk population. However, the death/hospitalization rate was influenced by risk factors in patient demographics and medical history. 
